---
title: "EDA of Endometrial Cancer Incidence in Sweden 1970-2024"
author: "Carmen Canedo"
output: html_notebook
---

```{r, warning=FALSE, include=FALSE}
library(tidyverse)
```

```{r}
ec <- read_delim("data/Statistikdatabasen_2026-01-28 21_04_12.csv",
               # File automatically comes with header. Could be worth deleting manually later
               skip = 1, delim = ";") 
  
```

```{r}
ec_long <- ec %>% 
  pivot_longer(cols = 2:56,
               names_to = "year",
               values_to = "count") %>% 
  rename(region = Region)
```


# Selection Overview
The `ec` table is pulled from [Socialstyrelsen Statistikdatabas för cancer](https://sdb.socialstyrelsen.se/if_can/val.aspx) with the following selections:

  + International incidence definition of morbidity IARC/IACR/ENCR
  + Tumor diagnosis (ICD-7): 172 Livmoderkropp (corpus uteri)
    - Note: ICD-10 is C54
  + Regardless of tumor type(s)
  + All regions, including Riket
  + Older age groups:
    - 0-17
    - 18-29
    - 30-44
    - 45-64
    - 65-79
    - 80+
  + Sex: women
  + Years: 1970-2024
  
# Summary Statistics
```{r}
ec_long %>% 
  summary()
```

```{r}
ec %>% 
  ggplot(aes(x = )) +
  geom_boxplot()
```

